---
title: "Regression Models Course Project"
date: "September 26, 2015"
output: pdf_document
---

## Executive Summary

Using data from the 1974 *Motor Trend* US cars report, 32 automobiles (1973-1974 models) were analyzed in an attempt to quantify the following two research questions:

1. Are automatic or manual transmissions are better for automobile mileage (MPG)?
2. What is the MPG difference between automobiles with automatic and manual transmissions?

Using stepwise least-squares regression to minimize the Akaike information criterion (AIC), transmission type was **not** found to be a statistically significant determinant in miles per gallon (MPG) prediction.  However this could be due in part to the small sample size used in the study (n=32).  That being said, the study indicates a **1.0 +/- 2.9 MPG difference** between manual and automatic transmissions.  The 95% confience interval for this differential is (-1.9, 3.9), and the estimator has a p-value of 0.4825, suggesting that the estimate is not statistically significant.


## Grading criteria:

* Did the student interpret the coefficients correctly?
* Did the student do some exploratory data analyses?
* Did the student fit multiple models and detail their strategy for model selection?
* Did the student answer the questions of interest or detail why the question(s) is (are) not answerable?
* Did the student do a residual plot and some diagnostics?
* Did the student quantify the uncertainty in their conclusions and/or perform an inference correctly?
* Was the report brief (about 2 pages long) for the main body of the report and no longer than 5 with supporting appendix of figures?
* Was the report done in Rmd (knitr)?


## 1. Introduction

Most automobiles run on gasoline (the exception being electric cars, biofuel-powered cars, and [DeLorian time machines](https://www.youtube.com/watch?v=3HYoq6vIVXc)).  People often look to the Miles Per Gallon (MPG) measurement to understand how efficenet their automobiles are at propelling themselves based on the car's fuel consumption.

It has been long rumored that automobiles with manual transmissions have higher MPGs than those with automatic transmissions.  We at *Motor Trent* wanted to investage this claim, and turned to our team of statisticians for answers.  Specifically, we wanted to investigate the following two research questions:

1. Are automatic or manual transmissions are better for automobile mileage (MPG)?
2. What is the MPG difference between automobiles with automatic and manual transmissions?

Using data from the 1974 *Motor Trend* US cars report (see the [mtcars](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html) database in the R [datasets](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/00Index.html) package), 32 automobiles from 1973 and 1974 were analyzed in an attempt to quantify our research questions.



## 2. Methods

According to the stepwise function, the best-fit regression equation for predicting miles per hour is

MPG = f(Weight + qsec + Automatic/Manual)



## 3. Results






```{r cars, eval=FALSE}
fit <- lm(mpg ~ ., data=mtcars)
step <- stepAIC(fit, direction="both")
step$anova
```

\pagebreak

## Appendix

A brief note on this file's formatting -- this file was developed using the knitr package in R; you can view the R Markdown file here: https://github.com/johnsug/dss/tree/master/07_regression/course_project.Rmd

```{r pressure, echo=FALSE}
plot(pressure)
```
