---
title: "Storm Tracker Analysis"
author: "github.com/johnsug"
date: "March 21, 2015"
output: ioslides_presentation
---

## Synopsis

This presentation is the result of my findings after exploring the [NOAA Storm Database](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2).

This presentation will answer:

- Which U.S. events are <font color="red">**most harmful**</font> with respect to population health?
- Which U.S. of events have the <font color="limegreen">**greatest economic**</font> consequences?

My analysis (and this presentation) was created using the [R](http://cran.r-project.org/) software suite in conjunction with [R Studio](http://www.rstudio.com/). 

## Data Processing

Loading data:

```{r dataProcessing, cache=TRUE}
## load data
d <- read.csv(bzfile("repdata-data-StormData.csv.bz2"))
```

## Data Processing, Cont'd

Query data to determine which events are <font color="red">**most harmful**</font> *(defined as fatalities plus injuries)*:

```{r processHarm, message=FALSE}
## query data
library(sqldf)
harm <- sqldf("select evtype, 
                sum(fatalities) as fatalities, 
                sum(injuries) as injuries 
              from d 
              group by evtype")

## harmful = fatalities + injuries
harm$total <- harm$fatalities + harm$injuries
harm <- sqldf("select * from harm 
              order by total desc 
              limit 10")
```

## Data Processing, Cont'd Again

Data munging (cleaning data):

```{r processEcon01}
## drop weird PROPDMGEXP values
`%notin%` <- function(x,y) !(x %in% y)  ## useful function
d <- d[d$PROPDMGEXP %notin% c('-','?','+',0:8,'h','H'),]
d$PROPDMGEXP[d$PROPDMGEXP=='m'] <- 'M'
d$PROPDMGEXP <- droplevels(d$PROPDMGEXP)

## drop weird CROPDMGEXP values
d <- d[d$CROPDMGEXP %notin% c('?',0,2),]
d$CROPDMGEXP[d$CROPDMGEXP=='k'] <- 'K'
d$CROPDMGEXP[d$CROPDMGEXP=='m'] <- 'M'
d$CROPDMGEXP <- droplevels(d$CROPDMGEXP)
```

## Data Processing, Cont'd Yet Again

More data munging (setting all figures to the same scale):

```{r processEcon02}
## re-scale property damage
d$property_cost[d$PROPDMGEXP=='K'] <- 
  d$PROPDMG[d$PROPDMGEXP=='K'] * 1000
d$property_cost[d$PROPDMGEXP=='M'] <- 
  d$PROPDMG[d$PROPDMGEXP=='M'] * 1000000
d$property_cost[d$PROPDMGEXP=='B'] <- 
  d$PROPDMG[d$PROPDMGEXP=='B']*1000000000

## re-scale crop damage
d$crop_cost[d$CROPDMGEXP=='K'] <- 
  d$CROPDMG[d$CROPDMGEXP=='K'] * 1000
d$crop_cost[d$CROPDMGEXP=='M'] <- 
  d$CROPDMG[d$CROPDMGEXP=='M'] * 1000000
d$crop_cost[d$CROPDMGEXP=='B'] <- 
  d$CROPDMG[d$CROPDMGEXP=='B'] * 1000000000
```

## Data Processing, Almost Done

Query data to determine which events have the <font color="limegreen">**greatest economic impact**</font> *(in terms of property plus crop damage)*:

```{r processEcon03, message=FALSE}
## economic cost = property cost + crop cost
d$property_cost[is.na(d$property_cost)] <- 0
d$crop_cost[is.na(d$crop_cost)] <- 0
d$economic_cost <- d$property_cost + d$crop_cost

## query and scale data (in billions)
library(sqldf)
econ <- sqldf("select evtype, 
                sum(economic_cost) / 1000000000 as cost 
              from d 
              group by evtype
              order by cost desc
              limit 10")
```

## Data Processing, One Last Slide

Generate top ten harmful and costly event histograms:

```{r histograms}
## generate most harmful events histogram ('h_hist')
library(ggplot2)
harm$EVTYPE <- factor(harm$EVTYPE, 
                      levels=rev(c(as.character(harm$EVTYPE))))
h_hist <- ggplot(harm, aes(x=EVTYPE, y=total)) + coord_flip() + 
            geom_histogram(stat="identity", fill="red", alpha=.5) + 
            labs(x="Event Type", y="Affected Lives", title="Top 10 Most Harmful Storms")

## generate most costly events histogram ('e_hist')
econ$EVTYPE <- factor(econ$EVTYPE, 
                      levels=rev(c(as.character(econ$EVTYPE))))
e_hist <- ggplot(econ, aes(x=EVTYPE, y=cost)) + coord_flip() + 
            geom_histogram(stat="identity", fill="green", alpha=.35) + 
            labs(x="Event Type", y="Cost (in Billions)", title="Top 10 Most Costly Storms")


```

## Results: <font color="red">Most Harmful</font> Events

```{r harmPlot}
h_hist
```


## Results: <font color="limegreen">Greatest Economic</font> Events
```{r econPlot}
e_hist
```

## Take Aways:

- <font color="limegreen"><b>Floods</b></font> are the <font color="limegreen">most expensive</font> U.S. storm
- <font color="red"><b>Tornados</b></font> are the <font color="red">most harmful</font> U.S. storm
- If you see flooding, tornados, **AND** sharks, just go home

<center><a href=http://www.syfy.com/sharknado3><img src=http://pixel.nymag.com/imgs/daily/vulture/2014/07/31/sharknado/weather%20lady.o.jpg/a_560x0.jpg></a></center>
